---
title: "Lab 3 - Coffee ratings"
subtitle: "STA 210 - Spring 2022"
footer:  "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
editor: visual
format: 
  revealjs: 
    theme: slides.scss
    transition: fade
    slide-number: true
execute:
  freeze: auto
---

# Model diagnostics

## The data

```{r}
df <- tibble(
  x = seq(1:100),
  y = 2*x + 20 + rnorm(100, sd = 10)
)

p_data <- ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  coord_cartesian(xlim = c(0, 200), ylim = c(0, 500))

p_data
```

## The data + an outlier

```{r}
df2 <- df %>%
  mutate(type = "data") %>%
  bind_rows(tibble(x = 40, y = 500, type = "outlier"))

p_outlier <- ggplot(df2, aes(x = x, y = y, color = type)) +
  geom_point(show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  geom_smooth(data = df, method = "lm", se = FALSE, color = "#8F2D5630") +
  scale_color_manual(values = c("black", "magenta")) +
  coord_cartesian(xlim = c(0, 200), ylim = c(0, 500)) +
  annotate("point", x = 40, y = 500, size = 3, color = "magenta", shape = "circle open")

p_outlier
```

## The data + influential point

```{r}
df3 <- df %>%
  mutate(type = "data") %>%
  bind_rows(tibble(x = 200, y = 5, type = "influential"))

p_influential <- ggplot(df3, aes(x = x, y = y, color = type)) +
  geom_point(show.legend = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  geom_smooth(data = df, method = "lm", se = FALSE, color = "#8F2D5630") +
  scale_color_manual(values = c("black", "magenta")) +
  coord_cartesian(xlim = c(0, 200), ylim = c(0, 500)) +
  annotate("point", x = 200, y = 5, size = 3, color = "magenta", shape = "circle open")

p_influential
```

## Influential point

An observation is **influential** if removing it substantially changes the coefficients of the regression model.

::: columns
::: {.column width="50%"}
```{r}
#| out.width: "100%"

p_data
```
:::

::: {.column width="50%"}
```{r}
#| out.width: "100%"

p_influential
```
:::
:::

## Influential points

-   Influential points have a large impact on the coefficients and standard errors used for inference

-   These points can sometimes be identified in a scatterplot if there is only one predictor variable, this is often not the case when there are multiple predictors

-   We will use measures to quantify an individual observation's influence on the regression model: **leverage**, **standardized residuals**, and **Cook's distance**

## Remember `augment()`?

```{r}
augment(df_fit$fit)
```

## Model diagnostics {.smaller}

Use the `augment()` function to output statistics that can be used to diagnose the model, along with the predicted values and residuals:

::: nonincremental
-   outcome and predictor variables in the model
-   **`.fitted`**: predicted values
-   **`.se.fit`**: standard errors of predicted values
-   **`.resid`**: residuals
-   **`.hat`**: leverage
-   **`.sigma`**: estimate of residual standard deviation when the corresponding observation is dropped from model
-   **`.cooksd`**: Cook's distance
-   **`.std.resid`**: standardized residuals
:::

