---
title: "SLR: Prediction + model evaluation"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Ã‡etinkaya-Rundel"
footer:  "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%"
)
```

# Welcome

## Announcements

-   New on the course website: [FAQ](/course-faq.html)
-   New communication tool: Slack
    -   Find the invite link in your inbox / on Sakai announcements
    -   Use #general for questions, #random for random ðŸ¤ª
    -   Use code formatting for for questions involving code (see Course FAQ for a demo video)
-   My office hours: All virtual for now, hope to move 1 hour / week to in person later in the semester

## Hybrid teaching {.smaller}

-   Lectures:
    -   In person as long as university says so (and I don't have COVID)
    -   If you can't be in class (and you're well enough to follow along), watch live (or the recording later) on [Panopto](https://duke.hosted.panopto.com/Panopto/Pages/Sessions/List.aspx#folderID=%22d6c1d58a-cb6d-4732-9d4b-ae0c011bf767%22)
    -   Watching live and have questions? Post on Slack!
    -   In class and see someone ask a question on Slack? Please raise it to me!
-   Labs:
    -   Not live streamed / recorded
    -   Lab 2 (next Monday) - individual
    -   Lab 3 onwards - in teams, if teammates are in isolation, set up team Zoom calls

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling
library(usdata)      # for the county_2019 dataset
library(scales)      # for pretty axis labels
library(glue)        # for constructing character strings

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

# Application exercise

::: appex
ðŸ“‹ [github.com/sta210-s22/ae-2-dcbikeshare](https://github.com/sta210-s22/?q=ae-2-dcbikeshare&type=all&language=&sort=)
:::

# Uninsurance and high school graduation rates in NC

## Data source

-   The data come from [`usdata::county_2019`](https://openintrostat.github.io/usdata/reference/county_2019.html)
-   These data have been compiled from the 2019 American Community Survey

```{r map-prep}
# data prop for mapping
dfips <- maps::county.fips %>%
  as_tibble() %>% 
  extract(polyname, c("region", "subregion"), "^([^,]+),([^,]+)$") %>%
  filter(region %in% c("north carolina", "new york"))

map_county_2019 <- map_data("county") %>%
  as_tibble() %>%
  filter(region %in% c("north carolina", "new york")) %>%
  left_join(dfips) %>%
  mutate(fips = if_else(
    subregion == "currituck" & region == "north carolina", 37053L, fips
  )) %>%
  left_join(county_2019, by = "fips")

map_county_2019_nc <- map_county_2019 %>%
  filter(state == "North Carolina")
```

## Uninsurance rate

```{r}
#| out.width: "100%"
ggplot(map_county_2019_nc, 
       aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = uninsured)) +
  scale_fill_viridis_c(option = "E", labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = NULL, y = NULL, fill = NULL,
    title = "Percent uninsured (2015 - 2019)",
    subtitle = "Civilian noninstitutionalized population in NC"
  ) +
  coord_quickmap(clip = "off") +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.2, 0.1),
  ) +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "durham"), aes(fill = uninsured), color = "white") +
  annotate("text", x = -78.7, y = 36.3, label = "Durham County\n(12%)", hjust = 0, size = 4, color = "white", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "swain"), aes(fill = uninsured), color = "black") +
  annotate("text", x = -83.4, y = 35.9, label = "Swain County\n(21.5%)", hjust = 1, size = 4, fontface = "bold") +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "gates"), aes(fill = uninsured), color = "black") +
  annotate("text", x = -76.9, y = 36.8, label = "Gates County\n(6.6%)", hjust = 0, size = 4, color = "black", fontface = "bold")
```

## High school graduation rate

```{r}
#| out.width: "100%"
map_county_2019 %>%
  filter(state == "North Carolina") %>%
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = hs_grad)) +
  scale_fill_viridis_c(option = "D", labels = label_percent(scale = 1, accuracy = 1)) +
  labs(x = NULL, y = NULL, fill = NULL,
    title = "Percent high school graduate (2015 - 2019)",
    subtitle = "25 and older population in NC") +
  coord_quickmap(clip = "off") +
  theme_void() +
  theme(
    legend.direction = "horizontal",
    legend.position = c(0.2, 0.1),
  ) +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "durham"), aes(fill = hs_grad), color = "white") +
  annotate("text", x = -78.7, y = 36.3, label = "Durham County\n(88.4%)", hjust = 0, size = 4, color = "white", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "tyrrell"), aes(fill = hs_grad), color = "black") +
  annotate("text", x = -76.4, y = 36.3, label = "Tyrrell\nCounty\n(74%)", hjust = 0, size = 4, color = "black", fontface = "bold") +
  geom_polygon(data = map_county_2019_nc %>% filter(subregion == "dare"), aes(fill = hs_grad), color = "black") +
  annotate("text", x = -75.9, y = 35.2, label = "Dare\nCounty\n(94.2%)", hjust = 0, size = 4, color = "black", fontface = "bold")
```

## Examining the relationship

-   The [NC Labor and Economic Analysis Division (LEAD)](https://www.nc.gov/agency/labor-and-economic-analysis-division), which "administers and collects data, conducts research, and publishes information on the state's economy, labor force, educational, and workforce-related issues".
-   Suppose that an analyst working for LEAD is interested in the relationship between uninsurance and high school graduation rates in NC counties.

. . .

::: question
What type of visualization should the analyst make to examine the relationship between these two variables?
:::

## Data prep

```{r}
#| echo: true

county_2019_nc <- county_2019 %>%
  as_tibble() %>%
  filter(state == "North Carolina") %>%
  select(name, hs_grad, uninsured)

county_2019_nc
```

## Uninsurance vs. HS graduation rates

```{r nc-uninsured-hsgrad-scatter}
#| code-fold: true
#| echo: true

ggplot(county_2019_nc,
       aes(x = hs_grad, y = uninsured)) +
  geom_point() +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Uninsurance vs. HS graduation rates",
    subtitle = "North Carolina counties, 2015 - 2019"
  ) +
  geom_point(data = county_2019_nc %>% filter(name == "Durham County"), aes(x = hs_grad, y = uninsured), shape = "circle open", color = "#8F2D56", size = 4, stroke = 2) +
  geom_text(data = county_2019_nc %>% filter(name == "Durham County"), aes(x = hs_grad, y = uninsured, label = name), color = "#8F2D56", fontface = "bold", nudge_y = 3, nudge_x = 2)
```

## Modeling the relationship

```{r nc-uninsured-hsgrad-scatter-line}
#| code-fold: true
#| echo: true
ggplot(county_2019_nc, aes(x = hs_grad, y = uninsured)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Uninsurance vs. HS graduation rates",
    subtitle = "North Carolina counties, 2015 - 2019"
  )
```

## Fitting the model

With `fit()`:

```{r}
#| echo: true

nc_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(uninsured ~ hs_grad, data = county_2019_nc)

tidy(nc_fit)
```

## Augmenting the data

With `augment()` to add columns for predicted values (`.fitted`), residuals (`.resid`), etc.:

```{r}
#| echo: true

nc_aug <- augment(nc_fit$fit)
nc_aug
```

## Visualizing the model I {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   **Black circles:** Observed values (`y = uninsured`)
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"

p_nc_aug_base <- ggplot(nc_aug, aes(x = hs_grad)) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(x = "High school graduate", y = "Uninsured")

p_nc_aug_base +
  geom_point(aes(y = uninsured))
```
:::
:::

## Visualizing the model II {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   **Pink solid line:** Least squares regression line
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"

p_nc_aug_base +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink")
```
:::
:::

## Visualizing the model III {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   Pink solid line: Least squares regression line
-   **Maroon triangles:** Predicted values (`y = .fitted`)
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"

p_nc_aug_base +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink") +
  geom_point(aes(y = .fitted), color = "maroon", shape = "triangle", size = 2)
```
:::
:::

## Visualizing the model IV {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Black circles: Observed values (`y = uninsured`)
-   Pink solid line: Least squares regression line
-   Maroon triangles: Predicted values (`y = .fitted`)
-   **Gray dashed lines:** Residuals
:::
:::

::: {.column width="75%"}
```{r}
#| out-width: "100%"

p_nc_aug_base +
  geom_segment(aes(xend = hs_grad, y = uninsured, yend = .fitted), size = 0.3, linetype = "dashed", color = "gray20") +
  geom_point(aes(y = uninsured)) +
  geom_smooth(aes(y = uninsured), method = "lm", se = FALSE, color = "pink") +
  geom_point(aes(y = .fitted), color = "maroon", shape = "triangle", size = 2)
```
:::
:::

## Evaluating the model fit

::: question
How can we evaluate whether the model for predicting uninsurance rate from high school graduation rate for NC counties is a good fit?
:::

# Model evaluation

## Two statistics {.smaller}

-   **R-squared**, $R^2$ : Percentage of variability in the outcome explained by the regression model (in the context of SLR, the predictor)

    $$
    R^2 = Cor(x,y)^2 = Cor(y, \hat{y})^2
    $$

-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)

    $$
    RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
    $$

. . .

::: question
What indicates a good model fit?
Higher or lower $R^2$?
Higher or lower RMSE?
:::

## R-squared {.smaller}

-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)

-   Unitless

-   Calculate with `rsq()`:

    ```{r}
    #| echo: true
    rsq(nc_aug, truth = uninsured, estimate = .fitted)
    ```

## Interpreting R-squared {.smaller}

```{r}
nc_fit_rsq <- round(glance(nc_fit)$r.squared * 100, 1)
```

::: poll
ðŸ—³ï¸ **Vote on Slack**

The $R^2$ of the model for predicting uninsurance rate from high school graduation rate for NC counties is `r nc_fit_rsq`%.
Which of the following is the correct interpretation of this value?

::: nonincremental
-   High school graduation rates correctly predict `r nc_fit_rsq`% of uninsurance rates in NC counties.
-   `r nc_fit_rsq`% of the variability in uninsurance rates in NC counties can be explained by high school graduation rates.
-   `r nc_fit_rsq`% of the variability in high school graduation rates in NC counties can be explained by uninsurance rates.
-   `r nc_fit_rsq`% of the time uninsurance rates in NC counties can be predicted by high school graduation rates.
:::
:::

## Alternative approach for R-squared

Alternatively, use `glance()` to construct a single row summary of the model fit, including $R^2$:

```{r}
#| echo: true

glance(nc_fit)
glance(nc_fit)$r.squared
```

## RMSE

-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)

-   Same units as the outcome variable

-   Calculate with `rmse()`:

    ```{r}
    #| echo: true

    rmse(nc_aug, truth = uninsured, estimate = .fitted)
    ```

-   The value of RMSE is not very meaningful on its own, but it's useful for comparing across models (more on this when we get to regression with multiple predictors)

## Obtaining R-squared and RMSE {.smaller}

-   Use `rsq()` and `rmse()`, respectively

    ```{r}
    #| echo: true
    #| eval: false

    rsq(nc_aug, truth = uninsured, estimate = .fitted)
    rmse(nc_aug, truth = uninsured, estimate = .fitted)
    ```

-   First argument: data frame containing `truth` and `estimate` columns

-   Second argument: name of the column containing `truth` (observed outcome)

-   Third argument: name of the column containing `estimate` (predicted outcome)

## Purpose of model evaluation

-   $R^2$ tells us how our model is doing to predict the data we *already have*
-   But generally we are interested in prediction for a new observation, not for one that is already in our sample, i.e. **out-of-sample prediction**
-   We have a couple ways of *simulating* out-of-sample prediction before actually getting new data to evaluate the performance of our models

# Splitting data

## Spending our data

-   There are several steps to create a useful model: parameter estimation, model selection, performance assessment, etc.
-   Doing all of this on the entire data we have available leaves us with no other data to assess our choices
-   We can allocate specific subsets of data for different tasks, as opposed to allocating the largest possible amount to the model parameter estimation only (which is what we've done so far)

## Simulation: data splitting {.smaller}

::: columns
::: {.column width="30%"}
::: nonincremental
-   Take a random sample of 10% of the data and set aside (testing data)
-   Fit a model on the remaining 90% of the data (training data)
-   Use the coefficients from this model to make predictions for the testing data
-   Repeat 10 times
:::
:::

::: {.column width="70%"}
```{r}
#| out.width: "100%"

set.seed(345)

n_folds <- 10

county_2019_nc_folds <- county_2019_nc %>%
  slice_sample(n = nrow(county_2019_nc)) %>%
  mutate(fold = rep(1:n_folds, n_folds)) %>%
  arrange(fold)

predict_folds <- function(i) {
  fit <- lm(uninsured ~ hs_grad, data = county_2019_nc_folds %>% filter(fold != i))
  predict(fit, newdata = county_2019_nc_folds %>% filter(fold == i)) %>%
    bind_cols(county_2019_nc_folds %>% filter(fold == i), .fitted = .)
}

nc_fits <- map_df(1:n_folds, predict_folds)

p_nc_fits <- ggplot(nc_fits, aes(x = hs_grad, y = .fitted, group = fold)) +
  geom_line(stat = "smooth", method = "lm", se = FALSE, size = 0.3, alpha = 0.5) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Predicted uninsurance rate in NC",
    subtitle = glue("For {n_folds} different testing datasets")
    )

p_nc_fits
```
:::
:::

## Predictive performance {.smaller}

::: columns
::: {.column width="25%"}
::: question
::: nonincremental
-   How consistent are the predictions for different testing datasets?
-   How consistent are the predictions for counties with high school graduation rates in the middle of the plot vs. in the edges?
:::
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"

p_nc_fits
```
:::
:::

# Bootstrapping

## Bootstrapping our data {.smaller}

-   The idea behind bootstrapping is that if a given observation exists in a sample, there may be more like it in the population
-   With bootstrapping, we simulate resampling from the population by resampling from the sample we observed
-   Bootstrap samples are the sampled *with replacement* from the original sample and same size as the original sample
    -   For example, if our sample consists of the observations {A, B, C}, bootstrap samples could be {A, A, B}, {A, C, A}, {B, C, C}, {A, B, C}, etc.

## Simulation: bootstrapping {.smaller}

::: columns
::: {.column width="25%"}
::: nonincremental
-   Take a bootstrap sample -- sample with replacement from the original data, same size as the original data
-   Fit model to the sample and make predictions for that sample
-   Repeat many times
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"

n_boot <- 100

predict_boots <- function(i){
  boot <- county_2019_nc %>%
    slice_sample(n = nrow(county_2019_nc), replace = TRUE) %>%
    mutate(boot_samp = i)
  fit <- lm(uninsured ~ hs_grad, data = boot)
  predict(fit) %>% bind_cols(boot, .fitted = .)
}

set.seed(1234)
county_2019_nc_boots <- map_df(1:n_boot, predict_boots)

p_nc_boots <- ggplot(county_2019_nc_boots, aes(x = hs_grad, y = .fitted, group = boot_samp)) +
  geom_line(stat = "smooth", method = "lm", se = FALSE, size = 0.3, alpha = 0.5) +
  scale_x_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 1)) +
  labs(
    x = "High school graduate", y = "Uninsured",
    title = "Predicted uninsurance rate in NC",
    subtitle = glue("For {n_boot} bootstrap samples")
    )

p_nc_boots
```
:::
:::

## Predictive performance {.smaller}

::: columns
::: {.column width="25%"}
::: question
::: nonincremental
-   How consistent are the predictions for different bootstrap datasets?
-   How consistent are the predictions for counties with high school graduation rates in the middle of the plot vs. in the edges?
:::
:::
:::

::: {.column width="75%"}
```{r}
#| out.width: "100%"

p_nc_boots
```
:::
:::
