---
title: "MultiLR: Prediction + inferential models"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
  echo: true
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

# Welcome

## Topics

::: nonincremental
-   Predictions
-   Model selection
-   Checking conditions
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(NHANES)
library(knitr)
library(patchwork)
library(colorblindr)
library(pROC)
library(Stat2Data)
library(nnet)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

## NHANES Data

::: nonincremental
-   [National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/index.htm) is conducted by the National Center for Health Statistics (NCHS).
-   The goal is to *"assess the health and nutritional status of adults and children in the United States".*
-   This survey includes an interview and a physical examination.
:::

## Variables

**Goal:** Use a person's age and whether they do regular physical activity to predict their self-reported health rating.

-   Outcome: `HealthGen`: Self-reported rating of participant's health in general.
    Excellent, Vgood, Good, Fair, or Poor.

-   Predictors:

    -   `Age`: Age at time of screening (in years). Participants 80 or older were recorded as 80.
    -   `PhysActive`: Participant does moderate to vigorous-intensity sports, fitness or recreational activities.

## The data

```{r}
nhanes_adult <- NHANES %>%
  filter(Age >= 18) %>%
  select(HealthGen, Age, PhysActive, Education) %>%
  drop_na() %>%
  mutate(obs_num = 1:n())
```

```{r}
glimpse(nhanes_adult)
```

## Model in R

```{r}
health_fit <- multinom_reg() %>%
  set_engine("nnet") %>%
  fit(HealthGen ~ Age + PhysActive, data = nhanes_adult)

health_fit <- repair_call(health_fit, data = nhanes_adult)
```

## Model summary

```{r}
tidy(health_fit) %>% print(n = 12)
```

# Predictions

## Calculating probabilities {.smaller}

-   For categories $2,\ldots,K$, the probability that the $i^{th}$ observation is in the $j^{th}$ category is

    $$
    \hat{\pi}_{ij} = \frac{e^{\hat{\beta}_{0j} + \hat{\beta}_{1j}x_{i1} + \dots + \hat{\beta}_{pj}x_{ip}}}{1 + \sum\limits_{k=2}^K e^{\hat{\beta}_{0k} + \hat{\beta}_{1k}x_{i1} + \dots \hat{\beta}_{pk}x_{ip}}}
    $$

-   For the baseline category, $k=1$, we calculate the probability $\hat{\pi}_{i1}$ as

    $$
    \hat{\pi}_{i1} = 1- \sum\limits_{k=2}^K \hat{\pi}_{ik}
    $$

## Predicted health rating {.smaller}

We can use our model to predict a person's perceived health rating given their age and whether they exercise.

```{r}
health_aug <- augment(health_fit, new_data = nhanes_adult)
health_aug
```

## Actual vs. predicted health rating

For each observation, the predicted perceived health rating is the category with the highest predicted probability.

```{r}
health_aug %>% select(contains("pred"))
```

## Confusion matrix

```{r}
health_conf <- health_aug %>% 
  count(HealthGen, .pred_class, .drop = FALSE) %>%
  pivot_wider(names_from = .pred_class, values_from = n)

health_conf
```

## Actual vs. predicted health rating

::: question
Why do you think no observations were predicted to have a rating of "Excellent", "Fair", or "Poor"?
:::

```{r}
#| echo: false
#| out-width: "100%"
#| fig-width: 10
#| layout-ncol: 2

ggplot(nhanes_adult, aes(x = HealthGen)) + 
  geom_bar() +
  labs(title = "Self-reported rating of overall health")

ggplot(health_aug, aes(x = HealthGen, fill = .pred_class)) + 
  geom_bar() +
  scale_fill_OkabeIto() +
  labs(title = "Predicted vs.\nself-reported rating of overall health") +
  theme(legend.position = c(0.8, 0.8))
```

# Model selection for inference

## Comparing nested models {.smaller}

-   Suppose there are two models:
    -   Reduced model includes predictors $x_1, \ldots, x_q$
    -   Full model includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$
-   We want to test the following hypotheses:
    -   $H_0: \beta_{q+1} = \dots = \beta_p = 0$
    -   $H_A: \text{ at least 1 }\beta_j \text{ is not } 0$
-   To do so, we will use the **drop-in-deviance test** (very similar to logistic regression)

## Add `Education` to the model? {.smaller}

-   We consider adding the participants' `Education` level to the model.
    -   Education takes values `8thGrade`, `9-11thGrade`, `HighSchool`, `SomeCollege`, and `CollegeGrad`
-   Models we're testing:
    -   Reduced model: `Age`, `PhysActive`
    -   Full model: `Age`, `PhysActive`, `Education`

. . .

$$
\begin{align}
&H_0: \beta_{9-11thGrade} = \beta_{HighSchool} = \beta_{SomeCollege} = \beta_{CollegeGrad} = 0\\
&H_a: \text{ at least one }\beta_j \text{ is not equal to }0
\end{align}
$$

## Add `Education` to the model?

```{r}
reduced_fit <- multinom_reg() %>%
  set_engine("nnet") %>%
  fit(HealthGen ~ Age + PhysActive,
  data = nhanes_adult)

full_fit <- multinom_reg() %>%
  set_engine("nnet") %>%
  fit(HealthGen ~ Age + PhysActive + Education,
  data = nhanes_adult)
  
reduced_fit <- repair_call(reduced_fit, data = nhanes_adult)
full_fit <- repair_call(full_fit, data = nhanes_adult)
```

## Add `Education` to the model? {.smaller}

```{r}
anova(reduced_fit$fit, full_fit$fit, test = "Chisq") %>%
  kable(digits = 3)
```

. . .

At least one coefficient associated with `Education` is non-zero.
Therefore, we will include `Education` in the model.

## Model with `Education` {.smaller}

```{r}
tidy(full_fit, conf.int = T) %>% print(n = 28)
```

## Compare NHANES models using AIC

Reduced model:

```{r}
glance(reduced_fit)$AIC
```

. . .

Full model:

```{r}
glance(full_fit)$AIC
```

# Checking conditions for inference

## Conditions for inference

We want to check the following conditions for inference for the multinomial logistic regression model:

1.  Linearity: Is there a linear relationship between the log-odds and the predictor variables?

2.  Randomness: Was the sample randomly selected?
    Or can we reasonably treat it as random?

3.  Independence: Are the observations independent?

## Checking linearity

Similar to logistic regression, we will check linearity by examining empirical logit plots between each level of the response and the quantitative predictor variables.

```{r}
nhanes_adult <- nhanes_adult %>%
  mutate(
    Excellent = factor(if_else(HealthGen == "Excellent", "1", "0")),
    Vgood = factor(if_else(HealthGen == "Vgood", "1", "0")),
    Good = factor(if_else(HealthGen == "Good", "1", "0")),
    Fair = factor(if_else(HealthGen == "Fair", "1", "0")),
    Poor = factor(if_else(HealthGen == "Poor", "1", "0"))
  )
```

## Checking linearity

```{r}
#| layout-ncol: 2

emplogitplot1(Excellent ~ Age, data = nhanes_adult, 
              ngroups = 10, main = "Excellent vs. Age")
emplogitplot1(Vgood ~ Age, data = nhanes_adult, 
              ngroups = 10, main = "Vgood vs. Age")
```

## Checking linearity

```{r}
#| layout-ncol: 2
#| out-width: "100%"

emplogitplot1(Good ~ Age, data = nhanes_adult, 
              ngroups = 10, main = "Good vs. Age")
emplogitplot1(Fair ~ Age, data = nhanes_adult, 
              ngroups = 10, main = "Fair vs. Age")
```

## Checking linearity

```{r}
emplogitplot1(Poor ~ Age, data = nhanes_adult, 
              ngroups = 10, main = "Poor vs. Age")
```

. . .

`r emo::ji("white_check_mark")` The linearity condition is satisfied.
There is a linear relationship between the empirical logit and the quantitative predictor variable, Age.

## Checking randomness

We can check the randomness condition based on the context of the data and how the observations were collected.

-   Was the sample randomly selected?

-   If the sample was not randomly selected, ask whether there is reason to believe the observations in the sample differ systematically from the population of interest.

. . .

`r emo::ji("white_check_mark")` The randomness condition is satisfied.
We do not have reason to believe that the participants in this study differ systematically from adults in the U.S..

## Checking independence

We can check the independence condition based on the context of the data and how the observations were collected.

Independence is most often violated if the data were collected over time or there is a strong spatial relationship between the observations.

. . .

`r emo::ji("white_check_mark")` The independence condition is satisfied.
It is reasonable to conclude that the participants' health and behavior characteristics are independent of one another.

## Recap

-   Predictions
-   Model selection for inference
-   Checking conditions for inference
