---
title: "SLR: Model diagnostics"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

# Welcome

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling
library(openintro)   # for the duke_forest dataset
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(kableExtra)  # also for pretty tables
library(patchwork)   # arrange plots

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

# Mathematical models for inference

## The regression model, revisited

```{r}
#| echo: true

df_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(price ~ area, data = duke_forest)

tidy(df_fit) %>%
  kable(digits = 2)
```

## HT for the slope {.smaller}

**Hypotheses:** $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \ne 0$

. . .

**Test statistic:** Number of standard errors the estimate is away from the null

$$
T = \frac{\text{Estimate - Null}}{\text{Standard error}} \\
$$

. . .

**p-value:** Probability of observing a test statistic at least as extreme (in the direction of the alternative hypothesis) from the null value as the one observed

$$
p-value = P(|t| > |\text{test statistic}),
$$

calculated from a $t$ distribution with $n - 2$ degrees of freedom

## HT: Test statistic

```{r}
tidy(df_fit) %>%
  kable(digits = 2) %>%
  row_spec(2, background = "#D9E3E4")
```

$$
t = \frac{\hat{\beta}_1 - 0}{SE_{\hat{\beta}_1}} = \frac{159.48 - 0}{18.17} = 8.78
$$

## HT: p-value

::: columns
::: {.column width="70%"}
```{r}
tidy(df_fit) %>%
  kable(digits = 2) %>%
  row_spec(2, background = "#D9E3E4")
```
:::

::: {.column width="30%"}
```{r}
#| out.width: "100%"

normTail(L = -8.78, U = 8.78, df = nrow(duke_forest) - 2, xlim = c(-9,9), col = "#D9E3E4")
```
:::
:::

## Understanding the p-value

|  Magnitude of p-value   |            Interpretation             |
|:-----------------------:|:-------------------------------------:|
|     p-value \< 0.01     |     strong evidence against $H_0$     |
| 0.01 \< p-value \< 0.05 |    moderate evidence against $H_0$    |
| 0.05 \< p-value \< 0.1  |      weak evidence against $H_0$      |
|     p-value \> 0.1      | effectively no evidence against $H_0$ |

::: callout-important
These are general guidelines.
The strength of evidence depends on the context of the problem.
:::

## HT: Conclusion, in context

```{r}
tidy(df_fit) %>%
  kable(digits = 2) %>%
  row_spec(2, background = "#D9E3E4")
```

-   The data provide convincing evidence that the population slope $\beta_1$ is different from 0.
-   The data provide convincing evidence of a linear relationship between area and price of houses in Duke Forest.

## CI for the slope

$$
\text{Estimate} \pm \text{ (critical value) } \times \text{SE}
$$

. . .

$$
\hat{\beta}_1 \pm t^* \times SE_{\hat{\beta}_1}
$$

where $t^*$ is calculated from a $t$ distribution with $n-2$ degrees of freedom

## CI: Critical value

::: columns
::: {.column width="60%"}
```{r}
#| echo: true

# confidence level: 95%
qt(0.975, df = nrow(duke_forest) - 2)

# confidence level: 90%
qt(0.95, df = nrow(duke_forest) - 2)

# confidence level: 99%
qt(0.995, df = nrow(duke_forest) - 2)
```
:::

::: {.column width="40%"}
```{r}
#| out.width: "100%"

normTail(M = c(-1.984984, 1.984984), df = nrow(duke_forest) - 2, col = "#D9E3E4")
text(x = 0, y = 0.04, labels = "95%", cex = 2, col = "#5B888C")
```
:::
:::

## 95% CI: Calculation

```{r}
tidy(df_fit) %>% 
  kable(digits = 2) %>%
  row_spec(2, background = "#D9E3E4")
```

$$\hat{\beta}_1 = 159.48 \hspace{15mm} t^* = 1.98 \hspace{15mm} SE_{\hat{\beta}_1} = 18.17$$

. . .

$$
159.48 \pm 1.98 \times 18.17 = (123.50, 195.46)
$$

## 95% CI: Computation {.smaller}

```{r}
#| echo: true

tidy(df_fit, conf.int = TRUE, conf.level = 0.95) %>% 
  kable(digits = 2)
```

# Model conditions

## Model conditions

1.  **Linearity:** There is a linear relationship between the response and predictor variables
2.  **Constant variance:** The variability of the errors is equal for all values of the predictor variable, i.e. the errors are homeoscedastic
3.  **Normality:** The errors follow a normal distribution
4.  **Independence:** The errors are independent from each other

## Linearity

✅ The residuals vs. fitted values plot should not show a random scatter of residuals (no distinguishable pattern or structure)

```{r res-vs-fit}
df_aug <- augment(df_fit$fit)

ggplot(df_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylim(-1000000, 1000000) +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  )
```

## Residuals vs. fitted values

```{r}
#| echo: true
#| ref.label: "res-vs-fit"
#| fig.show: "hide"
```

## Application exercise

::: appex
📋 [github.com/sta210-s22/ae-3-duke-forest](https://github.com/sta210-s22/?q=ae-3-duke-forest&type=all&language=&sort=)
:::

```{r}
countdown(minutes = 5, font_size = "2em")
```

## Non-linear relationships

```{r}
set.seed(1234)

n = 100

df <- tibble(
  x = -49:50,
  e_curved = rnorm(n, 0, 150),
  y_curved = x^2 + e_curved,
  e_slight_curve = sort(rbeta(n, 5, 1) * 200) + rnorm(n, 0, 5),
  y_slight_curve = x + e_slight_curve,
  x_fan = seq(0, 3.99, 4 / n),
  y_fan = c(rnorm(n / 8, 3, 1), rnorm(n / 8, 3.5, 2), rnorm(n / 8, 4, 2.5), rnorm(n / 8, 4.5, 3), rnorm(n / 4, 5, 4), rnorm((n / 4) + 2, 6, 5))
)
```

::: columns
::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| fig.asp: 1.2

p1 <- ggplot(df, aes(x = x, y = y_curved)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )

curved_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(y_curved ~ x, data = df)

curved_aug <- augment(curved_fit$fit)

p2 <- ggplot(curved_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  ) +
  ylim(-2000, 2000)

p1 / p2 +
  plot_annotation(title = "Obviously curved")
```
:::

::: {.column width="50%"}
```{r}
#| out.width: "100%"
#| fig.asp: 1.2

p1 <- ggplot(df, aes(x = x, y = y_slight_curve)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )

slight_curve_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(y_slight_curve ~ x, data = df)

slight_curve_aug <- augment(slight_curve_fit$fit)

p2 <- ggplot(slight_curve_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  )

p1 / p2 +
  plot_annotation(title = "Not so obviously curved")
```
:::
:::

## Constant variance

✅ The vertical spread of the residuals should be relatively constant across the plot

```{r}
#| ref.label: "res-vs-fit"
```

## Non-constant variance

::: columns
::: {.column width="50%"}
```{r}
#| out.width: "100%"

ggplot(df, aes(x = x_fan, y = y_fan)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#8F2D56") +
  labs(
    x = "X", y = "Y",
    title = "Observed data + model"
    )
```
:::

::: {.column width="50%"}
```{r}
#| out.width: "100%"

fan_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(y_fan ~ x_fan, data = df)

fan_aug <- augment(fan_fit$fit)

ggplot(fan_aug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Fitted value", y = "Residual",
    title = "Residuals vs. fitted values"
  ) +
  ylim(-15, 15)
```
:::
:::

## Normality

```{r}
ggplot(df_aug, aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), binwidth = 100000, color = "white") +
  geom_function(
    #geom = "line",
    fun = dnorm,
    args = list(
      mean = mean(df_aug$.resid), 
      sd = sd(df_aug$.resid)
      ),
    lwd = 2,
    col = "#8F2D5690"
  ) +
  labs(
    x = "Residual",
    y = "Density",
    title = "Histogram of residuals"
  )
```

## Independence

-   We can often check the independence assumption based on the context of the data and how the observations were collected

-   If the data were collected in a particular order, examine a scatterplot of the residuals versus order in which the data were collected

✅ Based on available information, the error for one house does not tell us anything about the error for another use

## Recap

Used residual plots to check conditions for SLR:

::: columns
::: {.column width="50%"}
::: nonincremental
-   Linearity
-   Constant variance
:::
:::

::: {.column width="50%"}
::: nonincremental
-   Normality
-   Independence
:::
:::
:::

. . .

::: question
Which of these conditions are required for fitting a SLR?
Which for simulation-based inference for the slope for an SLR?
Which for inference with mathematical models?
:::

```{r}
countdown(minutes = 3, font_size = "2em")
```
