---
title: "Cross validation"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
  echo: true
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

# Welcome

## Topics

::: nonincremental
-   Cross validation for model evaluation
-   Cross validation for model comparison
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)
library(schrute)
```

## Data & goal {.smaller}

::: nonincremental
-   Data: The data come from the [**shrute**](https://bradlindblad.github.io/schrute/) package, and has been transformed using instructions from [Lab 4](/labs/lab-4.html)
-   Goal: Predict `imdb_rating` from other variables in the dataset
:::

```{r}
#| echo: true

office_episodes <- read_csv(here::here("slides", "data/office_episodes.csv"))
office_episodes
```

# Modeling prep

## Split data into training and testing

```{r}
set.seed(123)
office_split <- initial_split(office_episodes)
office_train <- training(office_split)
office_test <- testing(office_split)
```

## Specify model

```{r}
office_spec <- linear_reg() %>%
  set_engine("lm")

office_spec
```

# Model 1

## From yesterday's lab

-   Create a recipe that uses the new variables we generated
-   Denotes `episode_name` as an ID variable and doesn't use `air_date` as a predictor
-   Create dummy variables for all nominal predictors
-   Remove all zero variance predictors

## Create recipe

```{r}
office_rec1 <- recipe(imdb_rating ~ ., data = office_train) %>%
  update_role(episode_name, new_role = "id") %>%
  step_rm(air_date) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

office_rec1
```

## Create workflow

```{r}
office_wflow1 <- workflow() %>%
  add_model(office_spec) %>%
  add_recipe(office_rec1)

office_wflow1
```

## Fit model to training data

. . .

*Actually, not so fast!*

# Cross validation

## Spending our data

-   We have already established that the idea of data spending where the test set was recommended for obtaining an unbiased estimate of performance.
-   However, we usually need to understand the effectiveness of the model *before using the test set*.
-   Typically we can't decide on *which* final model to take to the test set without making model assessments.
-   Remedy: Resampling to make model assessments on training data in a way that can generalize to new data.

## Resampling for model assessment

**Resampling is only conducted on the training set**.
The test set is not involved.
For each iteration of resampling, the data are partitioned into two subsamples:

-   The model is fit with the **analysis set**.
-   The model is evaluated with the **assessment set**.

## Resampling for model assessment

![](images/lec-14/resampling.svg){fig-align="center"}

<br>

Source: Kuhn and Silge.
[Tidy modeling with R](https://www.tmwr.org/).

## Analysis and assessment sets

-   Analysis set is analogous to training set.
-   Assessment set is analogous to test set.
-   The terms *analysis* and *assessment* avoids confusion with initial split of the data.
-   These data sets are mutually exclusive.

## Cross validation

More specifically, **v-fold cross validation** -- commonly used resampling technique:

-   Randomly split your **training** **data** into v partitions
-   Use 1 partition for assessment, and the remaining v-1 partitions for analysis
-   Repeat v times, updating which partition is used for assessment each time

. . .

Let's give an example where `v = 3`...

## Cross validation, step 1

Randomly split your **training** **data** into 3 partitions:

<br>

![](images/lec-14/three-CV.svg){fig-align="center"}

## Split data

```{r}
#| echo: true

set.seed(345)
folds <- vfold_cv(office_train, v = 3)
folds
```

## Cross validation, steps 2 and 3

::: nonincremental
-   Use 1 partition for assessment, and the remaining v-1 partitions for analysis
-   Repeat v times, updating which partition is used for assessment each time
:::

![](images/lec-14/three-CV-iter.svg){fig-align="center"}

## Fit resamples

```{r}
#| echo: true

set.seed(456)

office_fit_rs1 <- office_wflow1 %>%
  fit_resamples(folds)

office_fit_rs1
```

## Cross validation, now what?

-   We've fit a bunch of models
-   Now it's time to use them to collect metrics (e.g., R-squared, RMSE) on each model and use them to evaluate model fit and how it varies across folds

## Collect CV metrics

```{r}
collect_metrics(office_fit_rs1)
```

## Deeper look into CV metrics

```{r}
cv_metrics1 <- collect_metrics(office_fit_rs1, summarize = FALSE) 

cv_metrics1
```

## Better tabulation of CV metrics

```{r}
cv_metrics1 %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  pivot_wider(id_cols = id, names_from = .metric, values_from = .estimate) %>%
  kable(col.names = c("Fold", "RMSE", "R-squared"))
```

## How does RMSE compare to y? {.smaller}

Cross validation RMSE stats:

```{r}
cv_metrics1 %>%
  filter(.metric == "rmse") %>%
  summarise(
    min = min(.estimate),
    max = max(.estimate),
    mean = mean(.estimate),
    sd = sd(.estimate)
  )
```

Training data IMDB score stats:

```{r}
office_episodes %>%
  summarise(
    min = min(imdb_rating),
    max = max(imdb_rating),
    mean = mean(imdb_rating),
    sd = sd(imdb_rating)
  )
```

## Cross validation jargon

-   Referred to as v-fold or k-fold cross validation
-   Also commonly abbreviated as CV

## Cross validation, for reals

-   To illustrate how CV works, we used `v = 3`:

    ::: nonincremental
    -   Analysis sets are 2/3 of the training set
    -   Each assessment set is a distinct 1/3
    -   The final resampling estimate of performance averages each of the 3 replicates
    :::

-   This was useful for illustrative purposes, but `v = 3` is a poor choice in practice

-   Values of `v` are most often 5 or 10; we generally prefer 10-fold cross-validation as a default

# Application exercise

::: appex
📋 [github.com/sta210-s22/ae-6-the-office-cv](https://github.com/sta210-s22/ae-6-the-office-cv)
:::

## Recap

::: nonincremental
-   Cross validation for model evaluation
-   Cross validation for model comparison
:::
