---
title: "MLR: Inference"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
  echo: true
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

# Welcome

## Topics

-   Conduct a hypothesis test for $\beta_j$

-   Calculate a confidence interval for $\beta_j$

-   Inference pitfalls

## Computational setup

```{r}
#| warning: false

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)      # for tables
library(patchwork)  # for laying out plots
library(rms)        # for vif

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

## Data: `rail_trail` {.smaller}

::: nonincremental
-   The Pioneer Valley Planning Commission (PVPC) collected data for ninety days from April 5, 2005 to November 15, 2005.
-   Data collectors set up a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.
:::

```{r}
rail_trail <- read_csv(here::here("slides", "data/rail_trail.csv"))
rail_trail
```

Source: [Pioneer Valley Planning Commission](http://www.fvgreenway.org/pdfs/Northampton-Bikepath-Volume-Counts%20_05_LTA.pdf) via the **mosaicData** package.

## Variables {.smaller}

**Outcome**:

`volume` estimated number of trail users that day (number of breaks recorded)

. . .

**Predictors**

::: nonincremental
-   `hightemp` daily high temperature (in degrees Fahrenheit)
-   `avgtemp` average of daily low and daily high temperature (in degrees Fahrenheit)
-   `season` one of "Fall", "Spring", or "Summer"
-   `cloudcover` measure of cloud cover (in oktas)
-   `precip` measure of precipitation (in inches)
-   `day_type` one of "weekday" or "weekend"
:::

# Conduct a hypothesis test for $\beta_j$

## Review: Simple linear regression (SLR)

```{r}
ggplot(rail_trail, aes(x = hightemp, y = volume)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High temp (F)", y = "Number of riders")
```

## SLR model summary

```{r}
rt_slr_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(volume ~ hightemp, data = rail_trail)

tidy(rt_slr_fit)
```

## SLR hypothesis test {.smaller}

```{r}
#| echo: false

tidy(rt_slr_fit)
```

1.  **Set hypotheses:** $H_0: \beta_1 = 0$ and $H_A: \beta_1 \ne 0$

. . .

2.  **Calculate test statistic and p-value:** The test statistic is $t = 6.72$ with a degrees of freedom of 88, and a p-value \< 0.0001.

. . .

3.  **State the conclusion:** With a small p-value, we reject $H_0$. The data provide strong evidence that high temperature is a helpful predictor for the number of daily riders.

## Multiple linear regression

```{r}
rt_mlr_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(volume ~ hightemp + season, data = rail_trail)

tidy(rt_mlr_main_fit)
```

## MLR hypothesis test: `hightemp` {.smaller}

1.  **Set hypotheses:** $H_0: \beta_{hightemp} = 0$ and $H_A: \beta_{hightemp} \ne 0$, given `weekday` is in the model

. . .

2.  **Calculate test statistic and p-value:** The test statistic is $t = 6.43$ with a degrees of freedom of 86, and a p-value \< 0.0001.

. . .

3.  **State the conclusion:** With such a small p-value, the data provides strong evidence against $H_0$, i.e., the data provide strong evidence that high temperature for the day is a helpful predictor in a model that already contains whether the given day is a weekday as a predictor for number of daily riders.

## The model for `season = Spring` {.smaller}

```{r}
#| echo: false

tidy(rt_mlr_main_fit) %>% kable(digits = 2)
```

<br>

. . .

$$
\begin{aligned}
\widehat{volume} &= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times \texttt{seasonSpring} - 76.84 \times \texttt{seasonSummer} \\
&= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times 1 - 76.84 \times 0 \\
&= -120.10 + 7.54 \times \texttt{hightemp}
\end{aligned}
$$

## The model for `season = Summer` {.smaller}

```{r}
#| echo: false

tidy(rt_mlr_main_fit) %>% kable(digits = 2)
```

<br>

. . .

$$
\begin{aligned}
\widehat{volume} &= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times \texttt{seasonSpring} - 76.84 \times \texttt{seasonSummer} \\
&= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times 0 - 76.84 \times 1 \\
&= -202.07 + 7.54 \times \texttt{hightemp}
\end{aligned}
$$

## The model for `season = Fall` {.smaller}

```{r}
#| echo: false

tidy(rt_mlr_main_fit) %>% kable(digits = 2)
```

<br>

. . .

$$
\begin{aligned}
\widehat{volume} &= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times \texttt{seasonSpring} - 76.84 \times \texttt{seasonSummer} \\
&= -125.23 + 7.54 \times \texttt{hightemp} + 5.13 \times 0 - 76.84 \times 0 \\
&= -125.23 + 7.54 \times \texttt{hightemp}
\end{aligned}
$$

## The models

Same slope, different intercepts

-   `season = Spring`: $-120.10 + 7.54 \times \texttt{hightemp}$
-   `season = Summer`: $-202.07 + 7.54 \times \texttt{hightemp}$
-   `season = Fall`: $-125.23 + 7.54 \times \texttt{hightemp}$

## Application exercise

::: appex
📋 [github.com/sta210-s22/ae-8-rail-trail](https://github.com/sta210-s22/ae-8-rail-trail)
:::

::: columns
::: {.column width="30%"}
::: question
**Ex 1.** Recreate the following visualization in R based on the results of the model.
:::
:::

::: {.column width="70%"}
```{r}
#| label: main-viz
#| echo: false
#| out.width: "100%"

rt_mlr_main_aug <- augment(rt_mlr_main_fit$fit)

ggplot(rt_mlr_main_aug, aes(x = hightemp, y = volume)) +
  geom_point(aes(color = season), alpha = 0.5) +
  geom_smooth(aes(x = hightemp, y = .fitted, color = season), 
              method = "lm", se = FALSE) +
  labs(x = "High temp (F)", y = "Number of riders") +
  theme(legend.position = c(0.2, 0.8))
```
:::
:::

## Application exercise

::: appex
📋 [github.com/sta210-s22/ae-8-rail-trail](https://github.com/sta210-s22/ae-8-rail-trail)
:::

::: question
**Ex 2.** Add an interaction effect between `hightemp` and `season` and comment on the significance of the interaction predictors.
Time permitting, visualize the interaction model as well.
:::

```{r}
#| echo: false

countdown(minutes = 10)
```

# Confidence interval for $\beta_j$

## Confidence interval for $\beta_j$ {.smaller}

-   The $C%$ confidence interval for $\beta_j$ $$\hat{\beta}_j \pm t^* SE(\hat{\beta}_j)$$ where $t^*$ follows a $t$ distribution with $n - p - 1$ degrees of freedom.

-   Generically, we are $C%$ confident that the interval LB to UB contains the population coefficient of $x_j$.

-   In context, we are $C%$ confident that for every one unit increase in $x_j$, we expect $y$ to change by LB to UB units, holding all else constant.

## Confidence interval for $\beta_j$

```{r}
tidy(rt_mlr_main_fit, conf.int = TRUE)
```

## CI for `hightemp`

```{r}
#| echo: false

tidy(rt_mlr_main_fit, conf.int = TRUE)
```

<br>

We are 95% confident that for every degrees Fahrenheit the day is warmer, we expect the number of riders to increase by 5.21 to 9.87, on average, holding season constant.

## CI for `seasonSpring`

```{r}
#| echo: false

tidy(rt_mlr_main_fit, conf.int = TRUE)
```

<br>

We are 95% confident that the number of riders on a Spring day is, on average, lower by 63.1 to higher by 73.4 compared to a Fall day, holding high temperature for the day constant.

# Inference pitfalls

## Large sample sizes

::: callout-caution
If the sample size is large enough, the test will likely result in rejecting $H_0: \beta_j = 0$ even $x_j$ has a very small effect on $y$.

::: nonincremental
-   Consider the **practical significance** of the result not just the statistical significance.

-   Use the confidence interval to draw conclusions instead of relying only p-values.
:::
:::

## Small sample sizes

::: callout-caution
If the sample size is small, there may not be enough evidence to reject $H_0: \beta_j=0$.

::: nonincremental
-   When you fail to reject the null hypothesis, **DON'T** immediately conclude that the variable has no association with the response.

-   There may be a linear association that is just not strong enough to detect given your data, or there may be a non-linear association.
:::
:::
