---
title: "Multiple linear regression (MLR)"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

# Welcome

## Announcements {.smaller}

-   Lab 3:

    ::: nonincremental
    -   Any questions about lab / teamwork?
    -   Due Friday, Feb 4 at 5pm
    :::

-   Exam 1:

    ::: nonincremental
    -   Released Friday, Feb 4
    -   Multiple choice questions (mostly conceptual) + open-ended exercises (like lab + homework)
    -   Open book, open internet, open questions to me + Rick (head TA) only
    -   No communication with others or posting questions on the internet allowed
    -   What can you do to start preparing?
        -   Review readings, assignments, feedback returned

        -   Organize your notes

        -   Come to office hours with questions
    :::

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(tidymodels)  # for modeling
library(openintro)   # for the duke_forest dataset
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(patchwork)   # for laying out plots
library(GGally)      # for pairwise plots

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

# Considering multiple variables

## House prices in Levittown

-   The data set contains the sales price and characteristics of 85 homes in Levittown, NY that sold between June 2010 and May 2011.
-   Levittown was built right after WWII and was the first planned suburban community built using mass production techniques.
-   The article ["Levittown, the prototypical American suburb -- a history of cities in 50 buildings, day 25"](https://www.theguardian.com/cities/2015/apr/28/levittown-america-prototypical-suburb-history-cities) gives an overview of Levittown's controversial history.

## Analysis goals

-   We would like to use the characteristics of a house to understand variability in the sales price.

-   To do so, we will fit a **multiple linear regression model**.

-   Using our model, we can answers questions such as

    ::: nonincremental
    -   What is the relationship between the characteristics of a house in Levittown and its sale price?
    -   Given its characteristics, what is the expected sale price of a house in Levittown?
    :::

## The data {.smaller}

```{r}
#| echo: true

levittown <- read_csv(here::here("slides/data/homeprices.csv"))
levittown
```

## Variables {.smaller}

**Predictors**:

::: nonincremental
-   `bedrooms`: Number of bedrooms
-   `bathrooms`: Number of bathrooms
-   `living_area`: Total living area of the house (in square feet)
-   `lot_size`: Total area of the lot (in square feet)
-   `year_built`: Year the house was built
-   `property_tax`: Annual property taxes (in USD)
:::

**Response:** `sale_price`: Sales price (in USD)

## EDA: Response variable

```{r}
ggplot(data = levittown, aes(x = sale_price)) +
  geom_histogram(binwidth = 25000) +
  scale_x_continuous(labels = label_dollar()) +
  labs(
    title = "Sale prices of homes in Levittown",
    x = "Sale price (USD)",
    y = "Count"
  )
```

## EDA: Predictor variables

```{r}
#| out.width: "100%"
#| fig.width: 10

p1 <- ggplot(data = levittown, aes(x = bedrooms)) +
  geom_histogram() +
  scale_x_continuous(breaks = c(3, 4, 5))

p2 <- ggplot(data = levittown, aes(x = bathrooms)) +
  geom_histogram() +
  scale_x_continuous(breaks = 1:3)

p3 <- ggplot(data = levittown, aes(x = living_area)) +
  geom_histogram() +
  scale_x_continuous(breaks = c(1500, 2500))

p4 <- ggplot(data = levittown, aes(x = lot_size)) +
  geom_histogram()

p5 <- ggplot(data = levittown, aes(x = year_built)) +
  geom_histogram() +
  scale_x_continuous(breaks = c(1950, 1955, 1960))

p6 <- ggplot(data = levittown, aes(x = property_tax)) +
  geom_histogram() +
  scale_x_continuous(breaks = c(5000, 10000))

(p1 + p2 + p3) / (p4 + p5 + p6)
```

## EDA: Response vs. Predictors

```{r}
#| out.width: "100%"
#| fig.width: 10

p1 <- ggplot(data = levittown, aes(x = bedrooms, y = sale_price)) +
  geom_point() +
  scale_x_continuous(breaks = c(3, 4, 5))

p2 <- ggplot(data = levittown, aes(x = bathrooms, y = sale_price)) +
  geom_point() +
  scale_x_continuous(breaks = 1:3)

p3 <- ggplot(data = levittown, aes(x = living_area, y = sale_price)) +
  geom_point() +
  scale_x_continuous(breaks = c(1500, 2500))

p4 <- ggplot(data = levittown, aes(x = lot_size, y = sale_price)) +
  geom_point()

p5 <- ggplot(data = levittown, aes(x = year_built, y = sale_price)) +
  geom_point() +
  scale_x_continuous(breaks = c(1950, 1955, 1960))

p6 <- ggplot(data = levittown, aes(x = property_tax, y = sale_price)) +
  geom_point() +
  scale_x_continuous(breaks = c(5000, 10000))

(p1 + p2 + p3) / (p4 + p5 + p6)
```

## EDA: All variables

::: panel-tabset
## Plot

```{r}
#| ref.label: levittown-pairs
#| out.width: "80%"
#| fig.width: 12
```

## Code

```{r}
#| label: levittown-pairs
#| cache: true
#| echo: true
#| fig.show: hide
#| code-line-numbers: "|1"

ggpairs(levittown) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(angle = 45, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
```
:::

## Single vs. multiple predictors

So far we've used a *single predictor variable* to understand variation in a quantitative response variable

. . .

Now we want to use *multiple predictor variables* to understand variation in a quantitative response variable

# Multiple linear regression

## Multiple linear regression (MLR) {.smaller}

Based on the analysis goals, we will use a **multiple linear regression** model of the following form

$$
\begin{aligned}\hat{\text{sale_price}} ~ = & ~
\hat{\beta}_0 + \hat{\beta}_1 \text{bedrooms} + \hat{\beta}_2 \text{bathrooms} + \hat{\beta}_3 \text{living_area} \\
&+ \hat{\beta}_4 \text{lot_size} + \hat{\beta}_5 \text{year_built} + \hat{\beta}_6 \text{property_tax}\end{aligned}
$$

Similar to simple linear regression, this model assumes that at each combination of the predictor variables, the values `sale_price` follow a Normal distribution.

## Regression Model

**Recall:** The simple linear regression model assumes

$$
Y|X\sim N(\beta_0 + \beta_1 X, \sigma_{\epsilon}^2)
$$

. . .

**Similarly:** The multiple linear regression model assumes

$$
Y|X_1, X_2, \ldots, X_p \sim N(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p, \sigma_{\epsilon}^2)
$$

## The MLR model

For a given observation $(x_{i1}, x_{i2} \ldots, x_{ip}, y_i)$

$$
y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_{i} \hspace{8mm} \epsilon_i \sim N(0,\sigma_\epsilon^2)
$$

## Prediction

At any combination of the predictors, the mean value of the response $Y$, is

$$
\mu_{Y|X_1, \ldots, X_p} = \beta_0 + \beta_1 X_{1} + \beta_2 X_2 + \dots + \beta_p X_p
$$

. . .

Using multiple linear regression, we can estimate the mean response for any combination of predictors

$$
\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X_{1} + \hat{\beta}_2 X_2 + \dots + \hat{\beta}_p X_{p}
$$

## Model fit {.smaller}

```{r}
#| echo: true

price_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(sale_price ~ bedrooms + bathrooms + living_area + lot_size +
        year_built + property_tax, data = levittown)

tidy(price_fit) %>%
  kable(digits = 3)
```

## Model equation

$$
\begin{align}\hat{\text{price}} = & -7148818.957 - 12291.011 \times \text{bedrooms}\\[5pt]  
&+ 51699.236 \times \text{bathrooms}  + 65.903 \times \text{living area}\\[5pt]
&- 0.897 \times \text{lot size} +  3760.898 \times \text{year built}\\[5pt] 
&+ 1.476 \times \text{property tax}
\end{align}
$$

## Interpreting $\hat{\beta}_j$ {.smaller}

-   The estimated coefficient $\hat{\beta}_j$ is the expected change in the mean of $y$ when $x_j$ increases by one unit, <i>holding the values of all other predictor variables constant</i>.

. . .

-   **Example:** The estimated coefficient for `living_area` is 65.90. This means for each additional square foot of living area, we expect the sale price of a house in Levittown, NY to increase by \$65.90, on average, holding all other predictor variables constant.

## Prediction {.smaller}

::: question
What is the predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1,050 square feet of living area, 6,000 square foot lot size, built in 1948 with \$6,306 in property taxes?
:::

<br>

```{r}
#| echo: true

-7148818.957 - 12291.011 * 3 + 51699.236 * 1 + 
  65.903 * 1050 - 0.897 * 6000 + 3760.898 * 1948 + 
  1.476 * 6306
```

. . .

The predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes is **\$265,360**.

## Prediction, revisit

Just like with simple linear regression, we can use the `predict()` function in R to calculate the appropriate intervals for our predicted values:

```{r}
#| echo: true

new_house <- tibble(
  bedrooms = 3, bathrooms = 1, 
  living_area = 1050, lot_size = 6000, 
  year_built = 1948, property_tax = 6306
  )

predict(price_fit, new_house)
```

## Confidence interval for $\hat{\mu}_y$

::: question
Calculate a 95% confidence interval for the **estimated mean price** of houses in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes.
:::

<br>

```{r}
#| echo: true

predict(price_fit, new_house, type = "conf_int", level = 0.95)
```

## Prediction interval for $\hat{y}$

::: question
Calculate a 95% prediction interval for an individual house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with \$6306 in property taxes.
:::

<br>

```{r}
#| echo: true

predict(price_fit, new_house, type = "pred_int", level = 0.95)
```

## Cautions

-   **Do not extrapolate!** Because there are multiple predictor variables, there is the potential to extrapolate in many directions
-   The multiple regression model only shows **association, not causality**
    -   To show causality, you must have a carefully designed experiment or carefully account for confounding variables in an observational study

## Recap

::: nonincremental
-   Introduced multiple linear regression

-   Interpreted a coefficient $\hat{\beta}_j$

-   Used the model to calculate predicted values and the corresponding intervals
:::
