---
title: "MultiLR: Predictive models (cont.)"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    highlight-style: ayu-mirage
code-link: true
editor: visual
execute:
  freeze: auto
  echo: true
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

# Welcome

## Topics

::: nonincremental
-   Unbalanced data
-   Choosing the "final" model
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)
library(colorblindr)
library(themis)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

# From last time...

## Volcanoes

The data come from [The Smithsonian Institution](https://volcano.si.edu/), via [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-12/readme.md).

```{r}
#| warning: false

volcano <- read_csv(here::here("slides", "data/volcano.csv"))
names(volcano)
```

## Data prep

```{r}
volcano <- volcano %>%
  mutate(
    volcano_type = case_when(
      str_detect(primary_volcano_type, "Stratovolcano") ~ "Stratovolcano",
      str_detect(primary_volcano_type, "Shield") ~ "Shield",
      TRUE ~ "Other"
    ),
    volcano_type = fct_relevel(volcano_type, "Stratovolcano", "Shield", "Other")
  ) %>%
  select(
    volcano_type, latitude, longitude, 
    elevation, tectonic_settings, major_rock_1
    ) %>%
  mutate(across(where(is.character), as_factor))
```

## Split into testing/training

```{r}
set.seed(1234)

volcano_split <- initial_split(volcano)
volcano_train <- training(volcano_split)
volcano_test  <- testing(volcano_split)
```

## Specify a model

```{r}
volcano_spec <- multinom_reg() %>%
  set_engine("nnet")

volcano_spec
```

## Create cross validation folds

```{r}
set.seed(9876)

volcano_folds <- vfold_cv(volcano_train, v = 5)
volcano_folds
```

# Unbalanced data

## Unbalanced data

Remember that the observed volcano types are unbalanced:

```{r}
volcano %>% 
  count(volcano_type)
```

## Addressing unbalance {.smaller}

To address class unbalance, we generally use

-   **oversampling** data from levels that are less prevalent in the data
    -   e.g., `step_smote()`: Uses a technique called ["**S**ynthetic **M**inority **O**ver-sampling **Te**chnique"](https://jair.org/index.php/jair/article/view/10302/24590) to generate new examples of the minority class using nearest neighbors of these cases.
-   **downsampling** data from levels that are more prevalent in the data
    -   e.g., `step_downsample()`: Removes rows of a data set to make the occurrence of levels in a specific factor level equal.

## New recipe - oversample

```{r}
#| code-line-numbers: "|7"

volcano_rec3 <- recipe(volcano_type ~ ., data = volcano_train) %>%
  step_other(tectonic_settings) %>%
  step_other(major_rock_1) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_center(all_predictors()) %>%
  step_smote(volcano_type)
```

## New recipe - downsample

```{r}
#| code-line-numbers: "|7"

volcano_rec4 <- recipe(volcano_type ~ ., data = volcano_train) %>%
  step_other(tectonic_settings) %>%
  step_other(major_rock_1) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_center(all_predictors()) %>%
  step_downsample(volcano_type)
```

## New workflows

```{r}
volcano_wflow3 <- workflow() %>%
  add_recipe(volcano_rec3) %>%
  add_model(volcano_spec)

volcano_wflow4 <- workflow() %>%
  add_recipe(volcano_rec4) %>%
  add_model(volcano_spec)
```

## Fit resamples

```{r}
volcano_fit_rs3 <- volcano_wflow3 %>%
  fit_resamples(
    volcano_folds, 
    control = control_resamples(save_pred = TRUE)
    )

volcano_fit_rs4 <- volcano_wflow4 %>%
  fit_resamples(
    volcano_folds, 
    control = control_resamples(save_pred = TRUE)
    )
```

## Collect metrics

```{r}
collect_metrics(volcano_fit_rs3)
collect_metrics(volcano_fit_rs4)
```

## ROC curves - oversampling

```{r}
volcano_fit_rs3 %>%
  collect_predictions() %>%
  group_by(id) %>%
  roc_curve(
    truth = volcano_type,
    .pred_Stratovolcano:.pred_Other
  ) %>%
  autoplot()
```

## ROC curves - downsampling

```{r}
volcano_fit_rs4 %>%
  collect_predictions() %>%
  group_by(id) %>%
  roc_curve(
    truth = volcano_type,
    .pred_Stratovolcano:.pred_Other
  ) %>%
  autoplot()
```

## Addressing unbalance

::: question
Can you think of any issues resulting from over/down sampling?
:::

# Final model

## The "chosen" model

Let's stick to the models without over/down sampling.

From the application exercise:

```{r}
volcano_rec2 <- recipe(volcano_type ~ ., data = volcano_train) %>%
  step_other(tectonic_settings) %>%
  step_other(major_rock_1) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_center(all_predictors())

volcano_wflow2 <- workflow() %>%
  add_recipe(volcano_rec2) %>%
  add_model(volcano_spec)
```

## Fitting the final model

```{r}
#| code-line-numbers: "|1-4|6"

final_fit <- last_fit(
  volcano_wflow2, 
  split = volcano_split
  )

collect_metrics(final_fit)
```

## Confusion matrix

```{r}
collect_predictions(final_fit) %>%
  conf_mat(volcano_type, .pred_class)
```

## Confusion matrix - visualized

```{r}
collect_predictions(final_fit) %>%
  conf_mat(volcano_type, .pred_class) %>%
  autoplot()
```

## ROC curve

```{r}
collect_predictions(final_fit) %>%
  roc_curve(truth = volcano_type, .pred_Stratovolcano:.pred_Other) %>%
  autoplot()
```

## ROC curve - altogether

::: appex
📋 [github.com/sta210-s22/ae-11-volcanoes](https://github.com/sta210-s22/ae-11-volcanoes) - Exercise 3
:::

```{r}
#| echo: false

collect_predictions(final_fit) %>%
  roc_curve(truth = volcano_type, .pred_Stratovolcano:.pred_Other) %>%
  ggplot(aes(1 - specificity, sensitivity, color = .level)) +
  geom_abline(slope = 1, color = "gray50", lty = 2, alpha = 0.8) +
  geom_path(size = 1.5, alpha = 0.7) +
  labs(color = NULL) +
  scale_color_OkabeIto()
```

## Prediction

```{r}
final_fitted <- extract_workflow(final_fit)

new_volcano <- tibble(
  latitude = 35.9940,
  longitude = -78.8986,
  elevation = 404,
  tectonic_settings = "Subduction zone / Continental crust (>25 km)",
  major_rock_1 = "Andesite / Basaltic Andesite"
)

predict(
  final_fitted, 
  new_volcano, 
  type = "prob"
  )
```

## Acknowledgements

Inspired by

::: nonincremental
-   https://juliasilge.com/blog/multinomial-volcano-eruptions/
-   https://juliasilge.com/blog/nber-papers/
:::
