---
title: "Types of predictors"
subtitle: "STA 210 - Spring 2022"
author: "Dr. Mine Çetinkaya-Rundel"
footer: "[sta210-s22.github.io/website](https://sta210-s22.github.io/website/)"
logo: "images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

# Welcome

## Announcements

::: nonincremental
-   Congratulations on finishing Exam 1!
-   Grading of AEs
-   Questions on feedback vs. regrades
:::

## Topics

::: nonincremental
-   Mean-centering quantitative predictors

-   Using indicator variables for categorical predictors

-   Using interaction terms
:::

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(colorblindr)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

# Introduction

## Data: Peer-to-peer lender

Today's data is a sample of 50 loans made through a peer-to-peer lending club.
The data is in the `loan50` data frame in the **openintro** R package.

```{r}
loan50 %>%
  select(annual_income, debt_to_income, verified_income, interest_rate)
```

## Variables {.smaller}

**Predictors**:

::: nonincremental
-   `annual_income`: Annual income
-   `debt_to_income`: Debt-to-income ratio, i.e. the percentage of a borrower's total debt divided by their total income
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)
:::

**Outcome**: `interest_rate`: Interest rate for the loan

## Outcome: `interest_rate`

```{r}
ggplot(loan50, aes(x = interest_rate)) +
  geom_density() +
  labs(title = "Distribution of interest rate")
```

```{r}
loan50 %>%
  summarise(
    min = min(interest_rate),
    median = median(interest_rate),
    max = max(interest_rate)
  ) %>%
  kable()
```

## Predictors

```{r}
p1 <- ggplot(loan50, aes(y = verified_income)) +
  geom_bar() +
  labs(title = "Verified Income")

p2 <- ggplot(loan50, aes(x = debt_to_income)) +
  geom_histogram(binwidth = 0.25) +
  labs(title = "Debt to income ratio")

p3 <- ggplot(loan50, aes(x = annual_income)) +
  geom_histogram(binwidth = 20000) +
  labs(title = "Annual income")

p1 + p2 / p3
```

## Data manipulation 1: Rescale income

```{r}
#| echo: true

loan50 <- loan50 %>%
  mutate(annual_income_th = annual_income / 1000)

ggplot(loan50, aes(x = annual_income_th)) +
  geom_histogram(binwidth = 20) +
  labs(title = "Annual income (in $1000s)")
```

## Outcome vs. predictors

```{r}
p4 <- ggplot(loan50, aes(x = verified_income, y = interest_rate)) +
  geom_boxplot(fill = "#5B888C") +
  labs(
    y = "Interest Rate",
    x = "Income verification"
  )

p5 <- ggplot(loan50, aes(x = debt_to_income, y = interest_rate)) +
  geom_point(color = "#5B888C") +
  labs(
    y = NULL,
    x = "Debt to income ratio"
  )


p6 <- ggplot(loan50, aes(x = annual_income_th, y = interest_rate)) +
  geom_point(color = "#5B888C") +
  labs(
    y = NULL,
    x = "Annual income (in $1000s)"
  )

p4 + p5 / p6
```

## Fit regression model

```{r}
#| echo: true
int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(interest_rate ~ debt_to_income + verified_income  + annual_income_th,
      data = loan50)
```

## Summarize model results {.smaller}

```{r}
tidy(int_fit, conf.int = TRUE) %>%
  kable(digits = 3)
```

. . .

<br>

::: question
Describe the subset of borrowers who are expected to get an interest rate of 10.726% based on our model.
Is this interpretation meaningful?
Why or why not?
:::

# Mean-centered variables

## Mean-centering

If we are interested in interpreting the intercept, we can **mean-center** the quantitative predictors in the model.

We can mean-center a quantitative predictor $X_j$ using the following:

$$X_{j_{Cent}} = X_{j}- \bar{X}_{j}$$

. . .

If we mean-center all quantitative variables, then the intercept is interpreted as the expected value of the response variable when all quantitative variables are at their mean value.

## Data manipulation 2: Mean-center numeric predictors

```{r}
#| echo: true
loan50 <- loan50 %>%
  mutate(
    debt_inc_cent = debt_to_income - mean(debt_to_income), 
    annual_income_th_cent = annual_income_th - mean(annual_income_th)
    )
```

## Visualize mean-centered predictors

```{r}
p1 <- ggplot(loan50, aes(x = debt_to_income)) +
  geom_density(fill = "#5B888C", alpha = 0.7) +
  labs(x = NULL, title = "Debt to income ratio")

p2 <- ggplot(loan50, aes(x = debt_inc_cent)) +
  geom_density(fill = "#9dbdbf", alpha = 0.7) +
  labs(x = NULL, title = "Mean-centered\nDebt to income ratio")

p3 <- ggplot(loan50, aes(x = annual_income_th)) +
  geom_density(fill = "#5B888C", alpha = 0.7) +
  labs(x = NULL, title = "Annual income (in $1000s)")

p4 <- ggplot(loan50, aes(x = annual_income_th_cent)) +
  geom_density(fill = "#9dbdbf", alpha = 0.7) +
  labs(x = NULL, title = "Mean-centered\nAnnual income (in $1000s)")

(p1 + p3) / (p2 + p4)
```

## Using mean-centered variables in the model {.smaller}

::: question
How do you expect the model to change if we use the `debt_inc_cent` and `annual_income_cent` in the model?
:::

. . .

```{r}
int_cent_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(interest_rate ~ debt_inc_cent + verified_income + annual_income_th_cent, 
      data = loan50)

tidy(int_cent_fit, conf.int = T)
```

## Original vs. mean-centered model {.smaller}

::: columns
::: {.column width="50%"}
```{r}
tidy(int_fit) %>%
  select(term, estimate) %>%
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
tidy(int_cent_fit) %>%
  select(term, estimate) %>%
  kable(digits = 3)
```
:::
:::

# Indicator variables

## Indicator variables

-   Suppose there is a categorical variable with $K$ categories (levels)

-   We can make $K$ indicator variables - one indicator for each category

-   An **indicator variable** takes values 1 or 0

    -   1 if the observation belongs to that category
    -   0 if the observation does not belong to that category

## Data manipulation 3: Create indicator variables for `verified_income`

```{r}
#| echo: true

loan50 <- loan50 %>%
  mutate(
    not_verified = if_else(verified_income == "Not Verified", 1, 0),
    source_verified = if_else(verified_income == "Source Verified", 1, 0),
    verified = if_else(verified_income == "Verified", 1, 0)
  )
```

. . .

```{r}
loan50 %>%
  select(verified_income, not_verified, source_verified, verified) %>%
  slice(1, 3, 6)
```

## Indicators in the model {.smaller}

-   We will use $K-1$ of the indicator variables in the model.
-   The **baseline** is the category that doesn't have a term in the model.
-   The coefficients of the indicator variables in the model are interpreted as the expected change in the response compared to the baseline, holding all other variables constant.
-   This approach is also called **dummy coding**.

. . .

```{r}
loan50 %>%
  select(verified_income, source_verified, verified) %>%
  slice(1, 3, 6)
```

## Interpreting `verified_income` {.smaller}

```{r}
tidy(int_cent_fit, conf.int  = T) %>%
  kable(digits = 3) %>%
  row_spec(c(3,4), background = "#D9E3E4")
```

. . .

-   The baseline category is `Not verified`.
-   People with source verified income are expected to take a loan with an interest rate that is 2.211% higher, on average, than the rate on loans to those whose income is not verified, holding all else constant.
-   People with verified income are expected to take a loan with an interest rate that is 6.880% higher, on average, than the rate on loans to those whose income is not verified, holding all else constant.

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Interest rate vs. annual income

The lines are not parallel indicating there is an **interaction effect**.
The slope of annual income differs based on the income verification.

```{r}
p1 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Annual income (in $1000s)",
    y = "Interest rate"
  )

p2 <- ggplot(loan50, 
             aes(x = annual_income_th, y = interest_rate,
                 color = verified_income)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Annual income (in $1000s)", y = NULL, color = NULL) +
  theme(legend.position = c(0.6, 0.9)) +
  scale_color_OkabeIto()

p1 + p2 +
  plot_annotation(title = "Interest rate vs. annual income")
```

## Interaction term in model {.smaller}

```{r}
#| echo: true
int_cent_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(interest_rate ~ debt_inc_cent  +  debt_inc_cent + 
        annual_income_th_cent + verified_income * annual_income_th_cent,
      data = loan50)
```

```{r}
tidy(int_cent_int_fit) %>%
  kable(digits = 3) %>%
  row_spec(c(6,7), background = "#D9E3E4")
```

## Interpreting interaction terms

-   What the interaction means: The effect of annual income on the interest rate differs by -0.016 when the income is source verified compared to when it is not verified, holding all else constant.
-   Interpreting `annual_income` for source verified: If the income is source verified, we expect the interest rate to decrease by 0.023% (-0.007 + -0.016) for each additional thousand dollars in annual income, holding all else constant.

## Data manipulation 4: Create interaction variables {.smaller}

Defining the interaction variable in the model formula as `verified_income * annual_income_th_cent` is an implicit data manipulation step as well

```{r}
library(hardhat)

framed <- model_frame(interest_rate ~ debt_inc_cent  +  debt_inc_cent + annual_income_th_cent + verified_income * annual_income_th_cent, data = loan50)

model_matrix(framed$terms, framed$data) %>%
  glimpse()
```

# Wrap up

## Recap

-   Mean-centering quantitative predictors

-   Using indicator variables for categorical predictors

-   Using interaction terms

## Looking backward {.smaller}

Data manipulation, with **dplyr** (from **tidyverse**):

```{r}
#| echo: true
#| results: hide

loan50 %>%
  select(interest_rate, annual_income, debt_to_income, verified_income) %>%
  mutate(
    # 1. rescale income
    annual_income_th = annual_income / 1000,
    # 2. mean-center quantitative predictors
    debt_inc_cent = debt_to_income - mean(debt_to_income),
    annual_income_th_cent = annual_income_th - mean(annual_income_th),
    # 3. create dummy variables for verified_income
    source_verified = if_else(verified_income == "Source Verified", 1, 0),
    verified = if_else(verified_income == "Verified", 1, 0),
    # 4. create interaction variables
    `annual_income_th_cent:verified_incomeSource Verified` = annual_income_th_cent * source_verified,
    `annual_income_th_cent:verified_incomeVerified` = annual_income_th_cent * verified
  )
```

## Looking forward {.smaller}

**Feature engineering**, with **recipes** (from **tidymodels**):

```{r}
#| label: recipe
#| echo: true

loan_rec <- recipe( ~ ., data = loan50) %>%
  # 1. rescale income
  step_mutate(annual_income_th = annual_income / 1000) %>%
  # 2. mean-center quantitative predictors
  step_center(all_numeric_predictors()) %>%
  # 3. create dummy variables for verified_income
  step_dummy(verified_income) %>%
  # 4. create interaction variables
  step_interact(terms = ~ annual_income_th:verified_income)
```

## Recipe

```{r}
#| echo: true

loan_rec
```
