---
title: "AE 6: The Office"
subtitle: "Cross validation"
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-s22) and locate the repo titled `ae-6-the-office-cv-YOUR_GITHUB_USERNAME` to get started.
:::

## Packages

```{r}
#| label: load-pkgs
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)
```

## Load data

```{r}
#| label: load-data
#| message: false

office_episodes <- read_csv("data/office_episodes.csv")
```

## Split data into training and testing

Split your data into testing and training sets.

```{r}
#| label: initial-split

# add code here
```

## Specify model

Specify a linear regression model.
Call it `office_spec`.

```{r}
#| label: specify-model

# add code here
```

## Create recipe

Create the recipe from class.
Call it `office_rec1`.

```{r}
#| label: create-recipe

# add code here
```

## Create workflow

Create the workflow that brings together the model specification and recipe.
Call it `office_wflow1`.

```{r}
#| label: create-wflow

# add code here
```

## Cross validation

Conduct 10-fold cross validation.

```{r}
#| label: cv-tenfold

# add code here
```

## Summarize CV metrics

Summarize metrics from your CV resamples.

```{r}
#| label: cv-summarize

# add code here
```

## Another model - Model 2

Create a different (simpler, involving fewer variables) recipe and call it `office_rec2`.
Conduct 10-fold cross validation and summarize metrics.
Describe how the two models compare to each other based on cross validation metrics.
