---
title: "AE 7: Exam 2 Review"
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-s22) and locate the repo titled `ae-7-exam-2-review-YOUR_GITHUB_USERNAME` to get started.
:::

## Packages

```{r}
#| label: load-pkgs
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)
library(openintro)

# fix data!
loans_full_schema <- droplevels(loans_full_schema)
```

## Goal

Create a model for precicting `interest_rate`.

## View data

Note the dimensions of the data and the variable names.
Review the data dictionary.

```{r}
#| label: load-data

# add code here
```

## Split data into training and testing

Split your data into testing and training sets.

```{r}
#| label: initial-split

# add code here
```

## Write the model

Write the model for predicting interest rate (`interest_rate`) from debt to income ratio (`debt_to_income`), the term of loan (`term`), the number of inquiries (credit checks) into the applicant's credit during the last 12 months (`inquiries_last_12m`), whether there are any bankruptcies listed in the public record for this applicant (`bankrupt`), and the type of application (`application_type`).
The model should allow for the effect of to income ratio on interest rate to vary by application type.

*Add model here*

## Exploration

Explore characteristics of the variables you'll use for the model using the training data only.

```{r}
#| label: explore

# add code here
```

## Specify model

Specify a linear regression model.
Call it `office_spec`.

```{r}
#| label: specify-model

# add code here
```

## Create recipe

-   Predict `interest_rate` from `debt_to_income`, `term`, `inquiries_last_12m`, `public_record_bankrupt`, and `application_type`.
-   Mean center `debt_to_income`.
-   Make `term` a factor.
-   Create a new variable: `bankrupt` that takes on the value "no" if `public_record_bankrupt` is 0 and the value "yes" if `public_record_bankrupt` is 1 or higher. Then, remove `public_record_bankrupt`.
-   Interact `application_type` with `debt_to_income`.
-   Create dummy variables where needed and drop any zero variance variables.

```{r}
#| label: create-recipe

# add code here
```

## Create workflow

Create the workflow that brings together the model specification and recipe.

```{r}
#| label: create-wflow

# add code here
```

## Cross validation

Conduct 10-fold cross validation.

```{r}
#| label: cv-tenfold

# add code here
```

## Summarize CV metrics

Summarize metrics from your CV resamples.

```{r}
#| label: cv-summarize

# add code here
```

Why are we focusing on R-squared and RMSE instead of adjusted R-squared, AIC, BIC?

*\[Add response here\]*

## Next steps...

Depending on time, either

-   Create a workflow for another model with a new recipe (omitting the interaction variable), conduct CV, do model selection between these two, and then interpret the coefficients for the selected model.
-   Or interpret the coefficients for the one model you fit.

Make sure to interpret the intercept and slope coefficient for at least one numerical, one categorical, and one interaction predictor.
