---
title: "Log Transformations in Linear Regression"
execute: 
  echo: false
---

::: callout-note
The following supplemental notes were created by [Dr. Maria Tackett](https://www.mariatackett.net/) for STA 210.
They are provided for students who want to dive deeper into the mathematics behind regression and reflect some of the material covered in STA 211: Mathematics of Regression.
Additional supplemental notes will be added throughout the semester.
:::

```{r}
#| include: false

library(tidyverse)
library(knitr)
```

This document provides details about the model interpretation when the predictor and/or response variables are log-transformed.
For simplicity, we will discuss transformations for the simple linear regression model as shown in @eq-orig.

$$
\label{orig}
y = \beta_0 + \beta_1 x
$$ {#eq-orig}

All results and interpretations can be easily extended to transformations in multiple regression models.

Note: *log* refers to the natural logarithm.

## Log-transformation on the response variable

Suppose we fit a linear regression model with $\log(y)$, the log-transformed $y$, as the response variable.
Under this model, we assume a linear relationship exists between $x$ and $\log(y)$, such that $\log(y) \sim N(\beta_0 + \beta_1 x, \sigma^2)$ for some $\beta_0$, $\beta_1$ and $\sigma^2$.
In other words, we can model the relationship between $x$ and $\log(y)$ using the model in @eq-log-y.

$$
\log(y) = \beta_0 + \beta_1 x
$$ {#eq-log-y}

If we interpret the model in terms of $\log(y)$, then we can use the usual interpretations for slope and intercept.
When reporting results, however, it is best to give all interpretations in terms of the original response variable $y$, since interpretations using log-transformed variables are often more difficult to truly understand.

In order to get back on the original scale, we need to use the exponential function (also known as the anti-log), $\exp\{x\} = e^x$.
Therefore, we use the model in @eq-log-y for interpretations and predictions, we will use @eq-exp-y to state our conclusions in terms of $y$.

$$
\begin{aligned}
&\exp\{\log(y)\} = \exp\{\beta_0 + \beta_1 x\} \\[10pt]
\Rightarrow &y = \exp\{\beta_0 + \beta_1 x\} \\[10pt]
\Rightarrow &y = \exp\{\beta_0\}\exp\{\beta_1 x\}
\end{aligned}
$$ {#eq-exp-y}

In order to interpret the slope and intercept, we need to first understand the relationship between the mean, median and log transformations.

### Mean, Median, and Log Transformations

Suppose we have a dataset `y` that contains the following observations:

```{r}
y <- c(3,5,6,7,8)
y
```

If we log-transform the values of `y` then calculate the mean and median, we have

```{r}
log_y <- tibble(log_y = log(y))
summary <- log_y %>% 
  summarise(mean_log_y = mean(log_y), median_log_y = median(log_y))
kable(summary,digits=5)
```

If we calculate the mean and median of `y`, then log-transform the mean and median, we have

```{r}
centers <- tibble(y) %>% 
  summarise(mean_y = mean(y), median_y = median(y))
summary2 <- centers %>% 
  summarise(log_mean = log(mean_y), log_median = log(median_y))
kable(summary2, digits = 5)
```

This is a simple illustration to show

1.  $\text{Mean}[{\log(y)}] \neq \log[\text{Mean}(y)]$ - the mean and log are not commutable

2.  $\text{Median}[\log(y)] = \log[\text{Median}(y)]$ - the median and log are commutable

### Interpretaton of model coefficients

Using @eq-log-y, the mean $\log(y)$ for any given value of $x$ is $\beta_0 + \beta_1 x$; however, this does **not** indicate that the mean of $y = \exp\{\beta_0 + \beta_1 x\}$ (see previous section).
From the assumptions of linear regression, we assume that for any given value of $x$, the distribution of $\log(y)$ is Normal, and therefore symmetric.
Thus the median of $\log(y)$ is equal to the mean of $\log(y)$, i.e $\text{Median}(\log(y)) = \beta_0 + \beta_1 x$.

Since the log and the median are commutable, $\text{Median}(\log(y)) = \beta_0 + \beta_1 x \Rightarrow \text{Median}(y) = \exp\{\beta_0 + \beta_1 x\}$.
Thus, when we log-transform the response variable, the interpretation of the intercept and slope are in terms of the effect on the **median** of $y$.

**Intercept**: The intercept is expected median of $y$ when the predictor variable equals 0.
Therefore, when $x=0$,

$$
\begin{aligned}
&\log(y) = \beta_0 + \beta_1 \times 0 = \beta_0 \\[10pt]
\Rightarrow &y = \exp\{\beta_0\}
\end{aligned}
$$

Interpretation: When $x=0$, the median of $y$ is expected to be $\exp\{\beta_0\}$.

**Slope**: The slope is the expected change in the median of $y$ when $x$ increases by 1 unit.
The change in the median of $y$ is

$$
\exp\{[\beta_0 + \beta_1 (x+1)] - [\beta_0 + \beta_1 x]\} = \frac{\exp\{\beta_0 + \beta_1 (x+1)\}}{\exp\{\beta_0 + \beta_1 x\}} = \frac{\exp\{\beta_0\}\exp\{\beta_1 x\}\exp\{\beta_1\}}{\exp\{\beta_0\}\exp\{\beta_1 x\}} = \exp\{\beta_1\}
$$

Thus, the median of $y$ for $x+1$ is $\exp\{\beta_1\}$ times the median of $y$ for $x$.

Interpretation: When $x$ increases by one unit, the median of $y$ is expected to multiply by a factor of $\exp\{\beta_1\}$.

## Log-transformation on the predictor variable

Suppose we fit a linear regression model with $\log(x)$, the log-transformed $x$, as the predictor variable.
Under this model, we assume a linear relationship exists between $\log(x)$ and $y$, such that $y \sim N(\beta_0 + \beta_1 \log(x), \sigma^2)$ for some $\beta_0$, $\beta_1$ and $\sigma^2$.
In other words, we can model the relationship between $\log(x)$ and $y$ using the model in #eq-log-x.

$$
y = \beta_0 + \beta_1 \log(x)
$$ {#eq-log-x}

**Intercept**: The intercept is the mean of $y$ when $\log(x) = 0$, i.e. $x = 1$.

Interpretation: When $x = 1$ $(\log(x) = 0)$, the mean of $y$ is expected to be $\beta_0$.

**Slope**: The slope is interpreted in terms of the change in the mean of $y$ when $x$ is multiplied by a factor of $C$, since $\log(Cx) = \log(x) + \log(C)$.
Thus, when $x$ is multiplied by a factor of $C$, the change in the mean of $y$ is

$$
\begin{aligned}
[\beta_0 + \beta_1 \log(Cx)] - [\beta_0 + \beta_1 \log(x)] &= \beta_1 [\log(Cx) - \log(x)] \\[10pt] 
& = \beta_1[\log(C) + \log(x) - \log(x)] \\[10pt] 
& = \beta_1 \log(C)
\end{aligned}
$$

Thus the mean of $y$ changes by $\beta_1 \log(C)$ units.

Interpretation: When $x$ is multiplied by a factor of $C$, the mean of $y$ is expected to change by $\beta_1 \log(C)$ units.
For example, if $x$ is doubled, then the mean of $y$ is expected to change by $\beta_1 \log(2)$ units.

## Log-transformation on the the response and predictor variable

Suppose we fit a linear regression model with $\log(x)$, the log-transformed $x$, as the predictor variable and $\log(y)$, the log-transformed $y$, as the response variable.
Under this model, we assume a linear relationship exists between $\log(x)$ and $\log(y)$, such that $\log(y) \sim N(\beta_0 + \beta_1 \log(x), \sigma^2)$ for some $\beta_0$, $\beta_1$ and $\sigma^2$.
In other words, we can model the relationship between $\log(x)$ and $\log(y)$ using the model in @eq-log-x-y.

$$
\log(y) = \beta_0 + \beta_1 \log(x)
$$ {#eq-log-x-y}

Because the response variable is log-transformed, the interpretations on the original scale will be in terms of the median of $y$ (see the section on the log-transformed response variable for more detail).

**Intercept**: The intercept is the mean of $y$ when $\log(x) = 0$, i.e. $x = 1$.
Therefore, when $\log(x) = 0$,

$$
\begin{aligned}
&\log(y) = \beta_0 + \beta_1 \times 0 = \beta_0 \\[10pt]
\Rightarrow &y = \exp\{\beta_0\}
\end{aligned}
$$

Interpretation: When $x = 1$ $(\log(x) = 0)$, the median of $y$ is expected to be $\exp\{\beta_0\}$.

**Slope**: The slope is interpreted in terms of the change in the median $y$ when $x$ is multiplied by a factor of $C$, since $\log(Cx) = \log(x) + \log(C)$.
Thus, when $x$ is multiplied by a factor of $C$, the change in the median of $y$ is

$$
\begin{aligned}
\exp\{[\beta_0 + \beta_1 \log(Cx)] - [\beta_0 + \beta_1 \log(x)]\} &= 
\exp\{\beta_1 [\log(Cx) - \log(x)]\} \\[10pt] 
& = \exp\{\beta_1[\log(C) + \log(x) - \log(x)]\} \\[10pt] 
& = \exp\{\beta_1 \log(C)\} = C^{\beta_1}
\end{aligned}
$$

Thus, the median of $y$ for $Cx$ is $C^{\beta_1}$ times the median of $y$ for $x$.

Interpretation: When $x$ is multiplied by a factor of $C$, the median of $y$ is expected to multiple by a factor of $C^{\beta_1}$.
For example, if $x$ is doubled, then the median of $y$ is expected to multiply by $2^{\beta_1}$.
